Response to reviewers

### 2025-01-29 Manuscript/ecography_reponse_letter_2.tex
updates until table 2

We are grateful for the detailed and useful comments from all two reviewers
We have taken all their suggestions on board, integrated changes associated with the suggestion into the manuscript and responded to them in detail below
For clarity we have displayed the \textcolor{blue}{reviewers' comments in blue} and our own responses in black
We fully describe each change below.
We have also attached a tracked change version of the manuscript to hopefully improve clarifying our changes.

@@

\textcolor{blue}{Line 89-90: I think that some references are necessary here.}
We changed this sentence and merged it with the next one after discovering some previously missed literature:

\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. Novack-Gottshall (2016a,b))), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines 89-92. %@@@

@@

\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).} Lines 175-178. %@@@

@@

We remove the word ``say''.

@@

\textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines 124-124. %@@@

@@

We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
We added this footnote to Table 3:

@@

We believe that this specific point conflicts with reviewers 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
In this instance we thus prefer keeping the box 2 since we believe it might add value to at least some readers (and hopefully doesn't distract from the reading for other readers).
Furthermore, it is not entirely clear to us what the reviewer meant with the metrics relationship ``varying across spaces with different dimensionalities''.
Did they meant that spaces with different dimensionalities affect different metrics differently?
We did mention the effect of the curse in the methods section:

>>>>

We are grateful for the detailed and useful comments from all two reviewers
We have integrated changes associated with the suggestion into the manuscript and responded to them in detail below
For clarity we have displayed the \textcolor{blue}{reviewers' comments in blue} and our own responses in black
We fully describe each change below.
We have also attached a tracked change version of the manuscript to hopefully improve clarifying our changes.

@@

\textcolor{blue}{Line 89-90: I think that some references are necessary here.}

We changed this sentence and merged it with the next one after discovering some previously missed literature:

\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and processes (e.g. Novack-Gottshall (2016a,b))), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines 89-92. %@@@

@@

\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more common).} Lines 175-178. %@@@

@@

We removed the word ``say''.

@@

\textit{First, we simulated multiple independent random time dependent traits (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines 124-124. %@@@

@@

% TODO: Do you think we should briefly explain why we decided to stick with 5 axes and removed TPD, rather than drop to 4 like they suggest?
We have now removed all the curves for the \texttt{TPD} metrics for any space of more than four dimensions (including the empirical space).
We added this footnote to Table 3:

@@

We believe that this specific point conflicts with reviewer 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
In this instance we thus prefer keeping the box 2 since we believe it might add value to at least some readers (and hopefully doesn't distract from the reading for other readers).
Furthermore, it is not entirely clear to us what the reviewer meant with the metrics relationship ``varying across spaces with different dimensionalities''.
Did they meant that spaces with different dimensionalities affect different metrics differently?
We did mention the effect of the curse in the methods section:

### 2025-01-29 Manuscript/main_ecography_R2.tex
updates until table 2

Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors.
In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.
However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.
Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
For example competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
This emphasises the importance of not using a one-size-fits-all metric.
Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.

@@

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.

@@

Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and Table \ref{Tab:mechanisms}).
Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.

\begin{table}[htbp]
\caption{Glossary and equivalence of terms used in this papers; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
\centering
\begin{tabular}{p{2.5cm}p{4cm}p{5cm}p{5cm}}
Term in this paper & definition & in ecology & in macroevolution \\
\hline
Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, trait space, etc.  \\
\hline
Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs, etc. \\
\hline
Dimensions & Columns ($d$) & Traits, Ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
\hline
Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional characteristic*; e.g. geographical location & uni/multidimensional characteristic; e.g. landmark coordinates \\
\hline
Group & Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. \\
\hline
Metric & Statistic (i.e. a measure) & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index \\
\hline
Stressors & An algorithm removing a proportion of the observations & Introduction of invasive species, change of landscape use, climate change, pollution, experimental design, etc. & Mass extinction, Tectonic change, climate change, etc. \\
\hline
Stressors' intensity & The amount of observation removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction, amount of co-occurring species, etc. \\
\hline
\end{tabular}
* Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).
\end{table}

@@

First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).
Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).
These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
For example, for trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
These observations on which the traits are measured can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. (see Glossary in the supplementary materials, table 1).
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).

@@

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0} function in \texttt{R};][]{silverman1986density}.
Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is a factor increasing the scaling power of the algorithm (here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space (or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions).
We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
\end{itemize}

For the decreasing size, increasing density and shifting space stressors, we estimated the parameters $\rho$, $D$ and $\rho'$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
Note that the algorithm used here to simulate the stressors depends, to some extent, on the distribution of the data for each trait space.
The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just average increase in density for ``increasing density'' and just overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors).
These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\% across all metrics (i.e. any additional replicates beyond 171 added less that 1\% extra variance).
See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.

Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
However, if the data are distributed uniformly, this does not happen.
Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
However, this is not tackled here for both simplicity and to understand how they work in isolation. 
Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.

We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials figures 1 and 2).
In each pipeline, the dimensions where simulated as uncorrelated.
We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each observation on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with $200$ observations and $10$ dimensions generated in the exact same way, one might have a hypervolume of nearly $0$ and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

@@

  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}. & Metrics mainly capturing changes in richness (but less in divergence and not in regularity) \\

  Facilitation & Density change (``density'') & When observations with similar trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite. & Metrics capturing changes in regularity (but not richness and divergence) \\

  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable). & Metrics capturing changes in position (and richness) \\\\

  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar observations will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}). & Metrics capturing changes in divergence (but not richness and regularity) \\\\

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution), thus, the type of metric expected to recover the mechanism can be variable. These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

>>>>

Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to various stressors.
In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism.
However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.
Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
We show that different metrics perform differently when trying to capture signal from different approximations of common mechanisms relative to no mechanism at all (null).
For example, competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
This emphasises the importance of not using a one-size-fits-all metric.
Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.

@@

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
Usually this consists of: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space.
For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.

@@

Although some previous work has been done in understanding patterns in the context of mechanisms and processes (e.g. \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and Table \ref{Tab:mechanisms}).
Our goal is to test the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
We expect that our ability to capture a pattern does not only depend on the choice of metric (what) but also on the process and mechanism at hand (how and why).
To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based on extinction events on Hawaiian islands before and after the year 1500 (process; how), and how different metrics (pattern; what) can lead to different interpretations of the data.
We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data, with the exact same data sometimes being interpreted in opposite ways leading to sometimes opposite potential interpretations.

\begin{table}[htbp]
\caption{Glossary and equivalence of terms used in this paper; adapted from \cite{guillerme2018disprity} and \cite{mammola2021concepts}.}
\centering
\begin{tabular}{p{2.5cm}p{4cm}p{5cm}p{5cm}}
Term in this paper & definition & in ecology & in macroevolution \\
\hline
Trait space & Matrix ($n \times d$) with a structural relation between rows and columns & Functional space, morphospace , etc. & Morphospace, trait space, etc.  \\
\hline
Observations & Rows ($n$) & Taxa, field sites, environments, etc. & Taxa, specimen, populations, elements, OTUs, etc. \\
\hline
Dimensions & Columns ($d$) & Traits, ordination scores, distances, etc. & Traits, ordination scores, distances, etc. \\
\hline
Traits & Columns subset ($b \times n$; $b \leq d$) & uni/multidimensional characteristic*; e.g. body mass, geographical location & uni/multidimensional characteristic; e.g. landmark coordinates \\
\hline
Group & Rows subset ($m \times d$; $m \leq n$) & Treatments, phylogenetic group (clade), etc. & Clades, geological stratum, etc. \\
\hline
Metric & Statistic (i.e. a measure) & Dissimilarity index or metric, hypervolume, functional diversity, etc. & Disparity metric or index \\
\hline
Stressors & An algorithm removing a proportion of the observations & Introduction of invasive species, change of landscape use, climate change, pollution, experimental design, etc. & Mass extinction, Tectonic change, climate change, etc. \\
\hline
Stressors' intensity & The amount of observations removed (percentages) & Proportion of species affected, proportion of temperature change, etc. & Severity of mass extinction, amount of co-occurring species, etc. \\
\hline
\end{tabular}
\label{Tab:glossary}
* Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions expanded to include traits beyond the ones common in functional ecology (e.g. also in macroevolution, palaeontology, microbiology, etc.).
\end{table}

@@

First, we simulated multiple independent random time dependent traits (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).
Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive as traits).
These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
For example, for a trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
These observations on which the traits are measured can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. (see Table \ref{Tab:glossary}).
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more common).

@@

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0} function in \texttt{R};][]{silverman1986density}.
Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is a factor increasing the scaling power of the algorithm (here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space (or, inversely, observations are more likely to be ``preserved''/``surviving'' in less dense regions).
We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share trait combinations.
This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
\end{itemize}

For the decreasing size, increasing density and shifting space stressors, we estimated the parameters $\rho$, $D$ and $\rho'$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
Note that the algorithm used here to simulate the stressors depends, to some extent, on the distribution of the data for each trait space.
The algorithms used can lead to similar stressor effects depending on the simulated data distribution.
For example, ``increasing density'' and ``decreasing size'' stressors will lead to just what their names suggest if the observations are uniformly distributed (just an average increase in density for ``increasing density'' and just an overall decrease in size for the ``decreasing size'' stressors) but will have more or less similar effects (albeit not identical) in the case of a normal distribution (average increase in density and overall decrease in size for both ``increasing density'' and ``decreasing size'' stressors).
These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 intensities of data removal).
We used 171 replicates because that was empirically the smallest number of replicates required to reach a variance between replicates lower than 1\% across all metrics (i.e. any additional replicates beyond 171 added less than 1\% extra variance).
See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.

Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
However, if the data are distributed uniformly, this does not happen.
Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
However, this is not tackled here for both simplicity and to understand how the stressors work in isolation. 
Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautionary guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.

We repeated each simulation pipeline for 4 dimensional traits (generate a trait space and apply the stressor - Figure \ref{Fig:simulation_results}), and for 2 and 8 dimensional (see supplementary materials figures 1 and 2).
In each pipeline, the dimensions were simulated as uncorrelated.
We limited our simulations to a relatively small number of dimensions due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each observation on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with $200$ observations and $10$ dimensions generated in the exact same way, one might have a hypervolume of nearly $0$ and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

@@

  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}. & Metrics mainly capturing changes in richness* (but less in divergence* and not in regularity*) \\

  Facilitation & Density change (``density'') & When observations with similar trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite. & Metrics capturing changes in regularity* (but not richness* and divergence*) \\

  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable). & Metrics capturing changes in position (and richness*) \\\\

  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar observations will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}). & Metrics capturing changes in divergence* (but not richness* and regularity*) \\\\

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution), thus, the type of metric expected to recover the mechanism can be variable. These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.
    * Richness, divergence and regularity are dimensions of functional metrics that have been proposed to respectively reflect the sum of differences, the average differences and how regular the differences are among observations \citep{pavoine2011measuring,tucker2017guide,mammola2021concepts}.
}
}
    \label{Tab:mechanisms}
\end{table}
